---
title: "NOAA-Storm-Analysis"
output:
  pdf_document: default
  html_document:
    keep_md: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```


## Synopsis

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

The analysis undertaken addresses the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

In this analysis, initially, the total injuries and fatalities are calculated for each event type. Then, the top 10 most harmful events are shortlisted. Similarly, property damages and crop damages are computed for each event type and the top 10 events with the greatest economic consequences are shortlisted. Finally, the results are presented as plots and figures.


## Import Packages

```{r import packages}
library(dplyr)        # for data manipulation
library(ggplot2)      # for data visulization
library(grid)         # for grid graphics
library(scales)       # for scale functions
library(gtable)       # for grobs manipulation
```


## Data Processing

Check and download the **Storm data** into the **`data`** folder.

```{r download data}
# Check to see if the directory exists
if(!file.exists("./data")) {dir.create("data")}

## Check and download the data if not yet downloaded
if(!file.exists("./data/Storm_data.csv.bz2")) {
  url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url, destfile = "./data/Storm_data.csv.bz2")
}
```

Load the **Storm data** into a dataframe.

```{r load data, cache=TRUE}
df <- read.csv("./data/Storm_data.csv.bz2")
```

Look at the first few rows of the dataset.

```{r, cache=TRUE}
head(df)
```

Extract only the **required** columns into a new dataframe.

```{r new dataframe, cache=TRUE}
df_req <- df %>% 
          select(c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))
```

Look at the structure of the resulting dataset.

```{r}
str(df_req)
```

Look at the summary for the resulting dataset.

```{r}
summary(df_req)
```

Tabulate the unique values in **`PROPDMGEXP`** column along with their respective counts.

```{r}
table(df_req$PROPDMGEXP)
```

Tabulate the unique values in **`CROPDMGEXP`** column along with their respective counts.

```{r}
table(df_req$CROPDMGEXP)
```

From above, it can be observed that there are some erroneous values in the **`PROPDMGEXP`** and **`CROPDMGEXP`** columns such as **"?"**, **"+"**, etc. Since, their proportions are low, let us drop these observations.

```{r}
df_req <- df_req[!(df_req$PROPDMGEXP %in% c("-", "?", "+", "h", "H") | df_req$CROPDMGEXP == "?"), ]
```

Replace the **alphabetical exponents** with their respective **numerical equivalent values**, such as **'M'** by **'6'** (i.e. *Million*), etc.

```{r}
# Add the new factor levels
levels(df_req$PROPDMGEXP) <- c(levels(df_req$PROPDMGEXP), '0', '3', '6', '9')
levels(df_req$CROPDMGEXP) <- c(levels(df_req$CROPDMGEXP), '0', '3', '6', '9')
```

```{r}
# Replace the aplhabetical exponents from 'PROPDMGEXP'
df_req$PROPDMGEXP[(df_req$PROPDMGEXP %in% c("B", "b"))] <- '9'
df_req$PROPDMGEXP[(df_req$PROPDMGEXP %in% c("M", "m"))] <- '6'
df_req$PROPDMGEXP[(df_req$PROPDMGEXP %in% c("K", "k"))] <- '3'
```

```{r}
# Replace the aplhabetical exponents from 'CROPDMGEXP'
df_req$CROPDMGEXP[(df_req$CROPDMGEXP %in% c("B", "b"))] <- '9'
df_req$CROPDMGEXP[(df_req$CROPDMGEXP %in% c("M", "m"))] <- '6'
df_req$CROPDMGEXP[(df_req$CROPDMGEXP %in% c("K", "k"))] <- '3'
```

Also, replace the **'null'** factors by **'0'**.

```{r}
# Replace the 'null' exponent from 'PROPDMGEXP'
df_req$PROPDMGEXP[df_req$PROPDMGEXP == ""] <- '0'

# Replace the 'null' exponent from 'CROPDMGEXP'
df_req$CROPDMGEXP[df_req$CROPDMGEXP == ""] <- '0'
```

Convert the **'factor'** variables to **'numeric'** variables.

```{r}
# Convert 'PROPDMGEXP' to 'numeric'
df_req$PROPDMGEXP <- as.numeric(as.character(df_req$PROPDMGEXP))

# Convert 'CROPDMGEXP' to 'numeric'
df_req$CROPDMGEXP <- as.numeric(as.character(df_req$CROPDMGEXP))
```

Again, tabulate the unique values in **`PROPDMGEXP`** column along with their respective counts.

```{r}
table(df_req$PROPDMGEXP)
```

Similarly, tabulate the unique values in **`CROPDMGEXP`** column along with their respective counts.

```{r}
table(df_req$CROPDMGEXP)
```

Now, calculate the total **Property Damage** and **Crop Damage** and store them into the variables **`TOTPROPDMG`** and **`TOTCROPDMG`**, respectively. 

```{r}
# Total 'Property Damage'
df_req$TOTPROPDMG <- df_req$PROPDMG * (10 ^ df_req$PROPDMGEXP)

# Total 'Crop Damage'
df_req$TOTCROPDMG <- df_req$CROPDMG * (10 ^ df_req$CROPDMGEXP)
```

Calculate the **Total Damage** (i.e. sum of *Property Damage* and *Crop Damage*) and store it into the variable **`TOTDMG`**.

```{r}
df_req$TOTDMG <- df_req$TOTPROPDMG + df_req$TOTCROPDMG
```

Lastly,  convert the **`EVTYPE`** to **character** variable and convert to **upper** case.

```{r}
df_req$EVTYPE <- toupper(as.character(df_req$EVTYPE))
```


## Analysis

Look at the first few rows of the resulting dataset.

```{r, cache=TRUE}
head(df_req)
```

Group the dataset by the event type **`EVTYPE`** and calculate the **total** **`FATALITIES`**, **`INJURIES`**, **`TOTPROPDMG`**, **`TOTCROPDMG`** and **`TOTDMG`** for each event type. Also, store the results in a new dataframe.

```{r}
df_res <- df_req %>% 
          group_by(EVTYPE) %>% 
          summarise(across(c("FATALITIES", "INJURIES", "TOTPROPDMG", "TOTCROPDMG", "TOTDMG"), sum))
```

Extract the top 10 most harmful events with respect to **`FATALITIES`**.

```{r}
fatalities10 <- df_res %>% arrange(desc(FATALITIES)) %>% head(n=10)
```

Extract the top 10 most harmful events with respect to **`INJURIES`**.

```{r}
injuries10 <- df_res %>% arrange(desc(INJURIES)) %>% head(n=10)
```

Extract the top 10 events with greatest economic consequences with respect to **`TOTPROPDMG`**.

```{r}
propdmg10 <- df_res %>% arrange(desc(TOTPROPDMG)) %>% head(n=10)
```

Extract the top 10 events with greatest economic consequences with respect to **`TOTCROPDMG`**.

```{r}
cropdmg10 <- df_res %>% arrange(desc(TOTCROPDMG)) %>% head(n=10)
```

Extract the top 10 events with greatest economic consequences with respect to **`TOTDMG`**.

```{r}
totdmg10 <- df_res %>% arrange(desc(TOTDMG)) %>% head(n=10)
```


## Results

Look at the top 10 most harmful events with respect to population health.

```{r plot 1, fig.height=6, fig.width=7, cache=TRUE}
g1 <- fatalities10 %>% 
      ggplot(aes(FATALITIES, reorder(EVTYPE, FATALITIES))) + 
      geom_col() + 
      labs(title = "Top 10 most harmful events with respect to Fatalities", x = "Fatalities", y = "Event Type") +
      scale_x_continuous(limits = c(0, 6000), 
                         labels = label_number(accuracy = 1, scale = 1/1000, suffix = "K"))

g2 <- injuries10 %>% 
      ggplot(aes(INJURIES, reorder(EVTYPE, INJURIES))) + 
      geom_col() + 
      labs(title = "Top 10 most harmful events with respect to Injuries", x = "Injuries", y = "Event Type") + 
      scale_x_continuous(limits = c(0, 100000), 
                         labels = label_number(accuracy = 1, scale = 1/1000, suffix = "K"))

p1 <- ggplotGrob(g1)
p2 <- ggplotGrob(g2)
pl1 <- rbind(p1, p2, size = "first")
pl1$widths <- unit.pmax(p1$widths, p2$widths)
grid.newpage()
grid.draw(pl1)
```

- From above, it is observed that with respect to **_Fatalities_**, the most harmful event is the **Tornado** followed by **Excessive Heat** and **Flash Flood**.

- Again, with respect to **_Injuries_**, the most harmful event is the **Tornado**. But, here, it is followed by **Tstm Wind** and then **Flood**. Whereas, **Excessive Heat** has dropped down to the **4th** place and **Flash Flood** down to the **8th** place.

- There is an overlap of **7** events in the two lists.

- The total number of **_Injuries_** is much higher than the number of **_Fatalities_**.

Now, look at the top 10 events having the greatest economic consequences.

```{r plot 2, fig.height=9, fig.width=7, cache=TRUE}
g3 <- propdmg10 %>% 
      ggplot(aes(TOTPROPDMG, reorder(EVTYPE, TOTPROPDMG))) + 
      geom_col() + 
      labs(title = "Top 10 events having greatest economic consequences\nwith respect to Property Damage",
           x = "Property Damage", y = "Event Type") +
      scale_x_continuous(limits = c(0, 151000000000), 
                         labels = label_number(accuracy = 1, scale = 1/1000000000, suffix = "B"))

g4 <- cropdmg10 %>% 
      ggplot(aes(TOTCROPDMG, reorder(EVTYPE, TOTCROPDMG))) + 
      geom_col() + 
      labs(title = "Top 10 events having greatest economic consequences\nwith respect to Crop Damage",
           x = "Crop Damage", y = "Event Type") + 
      scale_x_continuous(limits = c(0, 15100000000), 
                         labels = label_number(accuracy = 1, scale = 1/1000000000, suffix = "B"))

g5 <- totdmg10 %>% 
      ggplot(aes(TOTDMG, reorder(EVTYPE, TOTDMG))) + 
      geom_col() + 
      labs(title = "Top 10 events having greatest economic consequences\nwith respect to Total Damage",
           x = "Total Damage", y = "Event Type") + 
      scale_x_continuous(limits = c(0, 151000000000), 
                         labels = label_number(accuracy = 1, scale = 1/1000000000, suffix = "B"))

p3 <- ggplotGrob(g3)
p4 <- ggplotGrob(g4)
p5 <- ggplotGrob(g5)
pl2 <- rbind(p3, p4, p5, size = "first")
pl2$widths <- unit.pmax(p3$widths, p4$widths, p5$widths)
grid.newpage()
grid.draw(pl2)
```

- From above, it can be observed that the two lists of top 10 events having greatest economic consequences with respect to the **_Property Damage_** and **_Crop Damage_** are considerably different.

- Since, the scale of **Property Damage** is much higher than that of **Crop Damage**, there is good overlap between the lists of top 10 events with respect to **_Property Damage_** and **_Total Damage_**, such as **Flood**, **Hurricane/Typhoon**, **Tornado**, etc.

- But, some other events such as **Drought**, **River Flood** and **Ice Storm** are introduced in the list of top 10 events with respect to **_Total Damage_** due to their high **Crop Damage** contribution. Whereas, other events such as **Tropical Storm**, **Winter Storm** and **High Wind** are eliminated.


## Conclusion

This analysis gives an overview of the types of events which are most harmful with respect to population health and the types of events which have the greatest economic consequences.

Different types of events have different effects, for example **Excessive Heat** causes high number of fatalities while **Storm Surge** results in great economic consequences.

But, there are also some key events, such as **Tornado** and **Flood** which are both greatly harmful to public health as well as have high economic consequence. And, hence, these are recommonded to be the top priority during preparation for disaster management and mitigation.

## End